Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-05-08 WSSSPE2_report/WSSSPE2_review.tex
Section 1: Wrap lines.
Sections 3 to 7:  touch up.

Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the 
United Kingdom. The SSI was founded to support the UK's research software 
community by cultivating better, more sustainable research software to enable
world-class research. Chue Hong's keynote started by making the point of
that the use of -- and reliance on -- software is pervasive in all areas of
world-leading research, showing examples from disciplines as diverse as
humanities and high-energy physics, and quoting Kersten Kleese Van Dam of the
Pacific Northwestern National Laboratory via a petition
campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}
at change.org, ``\emph{Today there are very few science areas left which do not
rely on IT and thus software for the majority of their research work. More
importantly key scientific advances in experimental and observational science
would have been impossible without better software.}'' He also cited Daniel
Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of
the NSF, ``\emph{Scientific discovery and innovation are advancing along
fundamentally new pathways opened by development of increasingly sophisticated
software. Software is an integral enabler of computation, experiment and theory,
and directly responsible for increased scientific productivity and enhancement
of researchers' capabilities.}''

Chue Hong drew attention to the issue that in the cyber-infrastructure and
high-performance community, we all too often disregard the hundreds of thousands
of researchers developing software thinking they are the long tail. In fact, the
numbers point to the fact that they are actually the mainstream and we
(referring to people including the audience of the talk) are the elite minority.
He emphasized that software is no longer special; it is both essential to and
common in scientific research. A 2014
survey\footnote{http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers}
conducted by the SSI polled researchers from 15 research-intensive UK
universities (406 respondents covering a representative range of funders,
disciplines, and seniority) reported that 92\% confirmed the use of research
software and 89\% affirmed that it would be impossible or difficult to conduct
research without software. Nevertheless the British research community is just
starting to understand the magnitude of the issue. Whilst many researchers make
use of software such as MATLAB, SPSS, and Excel, data from the aforementioned
SSI survey shows that over half (56\%) of respondents developed their own
research software (which equates to over 140,000 researchers if extrapolated
across the UK) and yet 71\% had no formal software-development training, having
to rely on their own coding skills.

Examining another aspect of the size of the research software community, 
Chue Hong noted that the costs of software-reliant research in the United Kingdom
included \textsterling 840 million of investment in the financial year
2013--2014, and this amount has risen by 3\% on average over the past four
years. About 30\% of total research investment has been spent on research that
relies on software over the last four financial years. These numbers stemmed
from an analysis by the SSI of data from $49,650$ grant titles and abstracts published on
Gateway to Research between years 2010 and 2014. A similar analysis of university 
jobs advertised in the same period discovered that despite this investment, only
4\% of positions were software development related, and of these only 17\% were
explicitly named as a software developer / software engineer position: the vast
majority being advertised as research associate / research assistant positions.
This in turn leads to the issue of career paths for those bridging the research
and software worlds, who are essential to support the use and further
development of research software, a point highlighted by a graphic showing UK
STEM graduate career paths\footnote{Source: The Scientific Century, Royal
Society, 2010 (revised to reflect first stage clarification from ``What Do PhD's
Do?'' study)} showing that only 3.5\% were able to secure permanent positions.

@@

\item determine three things they considered to be significant barriers to
software sustainability
\end{enumerate}
Once each group had come up with answers, all the answers were compiled, and the
attendees voted on which they thought were important by a show of hands.

@@

Subgroup B discussed four papers and had six participants. Because of the nature
of the papers, training emerged as a common theme, however the conversation was
wide-ranging, including incentives, reproducibility and funding to promote
sustainability.

@@

\subsubsection{Subgroup B Discussion}
The statement ``applied computer science is being attempted in academia without
any formal training'' kicked off our discussion. The group brought expertise in
several different training models, from a two-day Software Carpentry workshop,
to a three-week Open Science for Synthesis (OSS) program to semester-long
programs. We discussed the pros and cons of different training approaches, also
touching on informal learning, for example where people learning the necessary
skills because they had cross-disciplinary people (``boundary scientists'') in
their work environments of whom they could ask questions.

Some papers explored gaps in training of early career scientists. Industry
participants in our group confirm this. We asked ourselves, ``Are traditional
courses failing?'' We think yes. Changes to undergraduate curricula requirements
are difficult. But as programming models become more complex, we have to raise
the skill levels. Skills must be improved not only in traditional programming --
learning languages and algorithms -- but around professional software
development. We want to get to a point where ``Everyone has a new minimum now --
everyone knows Git.'' Developers also need training on licensing choices. Just
because a code is on Git does not mean that it is open.

``How do you get people to look for the training they need?'' wondered one
participant. People seek out opportunities like Software Carpentry to augment
skills. While some instruction can be done in an institutional curricula,
independent groups (non-profits, institutes) have more flexibility. Some asked
whether Silicon Valley be interested in funding training so that people are
better prepared to enter the workforce? Some felt that companies were reluctant
to deliver training because employees then leave the company. While others felt
that pushing all training to industry could lead to good technical people
``getting on the Google bus'' and leaving the sciences.

``How will we know when people are trained effectively in these new skills?'' We
discussed certification. It can be hard to build the recognition of Java or
database certifications amongst all technologies. OSS offers badges to those
completing training. However we need to demonstrate proficiency, not just
completion. Google Summer of Code is a big CV augmentation for
participants---can we duplicate something similar?

The group also discussed how to fund training. On participant observed that NYU
runs a six-week data-science training; companies grab the graduates and pay for
those they hire. OSS also used the NSF Software Institutes program as a vehicle
to fund training activities. Software Carpentry uses a collaborative teaching
approach where people publish open teaching materials and receive credit for
their use.

We then discussed career paths for those supporting sustainable software. While
tenure track is not the only option for students, the challenges for those who
remain in academia can be large. Research scientists are entirely dependent on
soft money, which can be unpredictable. Postdocs and those who do pursue tenure
track positions need to publish and see no rewards in software development.
These challenges were all identified at WSSSPE1. What actions would we recommend
to improve things? Altmetrics and download statistics may slowly change the
system and improving a developer's ability receive credit for time invested in
software development. NSF recognition of data products (including software) as
well as publications is also helping.

We asked ourselves, ``Where are there examples of things that are changing
because of this and how can we build momentum?'' One example demonstrates change
over time. In 2007, nanoHUB listed the academic reward structure as a problem in
an EDUCAUSE report where the authors note, ``In the future, nanoHUB researchers
are hoping to change the research culture. While they recognize that young
faculty members are unlikely to get tenure based on their nanoHUB contributions,
they hope to encourage faculty to think beyond their own research needs to
consider publishing tutorials and other Web content in their field on the
nanoHUB site.'' Fast forward seven years later where, in a 2014 iSGTW article,
quote nanoHUB Principal Investigator (PI) Gerhard Klimeck, ``A former student of
mine published eight tools on nanoHUB, serving over 6,000 people with his tools.
He then joined a university as a professor and introduced nanoHUB. Use of the
gateway from that university skyrocketed; he used nanoHUB in existing classes,
created new classes, and infused it in his research. Ultimately, the professor's
department head associates his two-year rise to tenure with the notoriety and
innovation he gained through nanoHUB.''

>>>>

Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the 
United Kingdom (UK). The SSI was founded to support the UK's research software 
community by cultivating better, more sustainable research software to enable
world-class research. Chue Hong's keynote started by making the point 
that the use of -- and reliance on -- software is pervasive in all areas of
world-leading research, showing examples from disciplines as diverse as
humanities and high-energy physics, quoting Kersten Kleese Van Dam of the
Pacific Northwestern National Laboratory via a petition
campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}
at change.org, ``\emph{Today there are very few science areas left which do not
rely on IT and thus software for the majority of their research work. More
importantly key scientific advances in experimental and observational science
would have been impossible without better software.}'' He also cited Daniel
Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of
the NSF, ``\emph{Scientific discovery and innovation are advancing along
fundamentally new pathways opened by development of increasingly sophisticated
software. Software is an integral enabler of computation, experiment and theory,
and directly responsible for increased scientific productivity and enhancement
of researchers' capabilities.}''

Chue Hong drew attention to the issue that in the cyber-infrastructure and
high-performance community, hundreds of thousands
of researchers developing software are all too often disregarded or considered the long tail. 
Actually, the numbers point to the fact that they are the mainstream. %and we
%(referring to people including the audience of the talk) are the elite minority.
He emphasized that software is no longer special; it is both essential to and
common in scientific research. A 2014
survey\footnote{http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers}
conducted by the SSI polled researchers from 15 research-intensive UK
universities (406 respondents covering a representative range of funders,
disciplines, and seniority). The survey reported that 92\% respondents confirmed the use of research
software and 89\% affirmed that it would be impossible or difficult to conduct
research without software. Nevertheless the British research community is just
starting to understand the magnitude of the issue. Whilst many researchers make
use of software such as MATLAB, SPSS, and Excel, data from the aforementioned
SSI survey shows that over half (56\%) of respondents developed their own
research software (which equates to over 140,000 researchers if extrapolated
across the UK) and yet 71\% of all UK researchers had no formal software-development training, having
to rely on their own coding skills.

Examining another aspect of the size of the research software community, 
Chue Hong noted that the costs of software-reliant research in the UK
included \textsterling 840 million of investment in the financial year
2013--2014, and this amount has risen by 3\% on average over the past four
years. About 30\% of total research investment has been spent on research that
relies on software over the last four financial years. These numbers stemmed
from an analysis by the SSI of data from $49,650$ grant titles and abstracts published on
Gateway to Research between years 2010 and 2014. A similar analysis of university 
jobs advertised in the same period discovered that despite this investment, only
4\% of positions were software development related, and of these only 17\% were
explicitly named as a software developer or software engineer positions: the vast
majority being advertised as research associate or research assistant positions.
This in turn leads to the issue of career paths for those bridging the research
and software worlds, who are essential to support the use and further
development of research software, a point highlighted by a graphic showing UK
STEM graduate career paths\footnote{Source: The Scientific Century, Royal
Society, 2010 (revised to reflect first stage clarification from ``What Do PhD's
Do?'' study)} showing that only 3.5\% were able to secure permanent positions.

@@

\item determine three things they considered to be significant barriers to
software sustainability.
\end{enumerate}
Once each group had come up with answers, all the answers were compiled, and the
attendees voted on which they thought were important by a show of hands.

@@

Subgroup B discussed four papers and had six participants. Because of the nature
of the papers, training emerged as a common theme. However the conversation was
wide-ranging, including incentives, reproducibility, and funding to promote
sustainability.

@@

\subsubsection{Subgroup B Discussion}
The statement ``applied computer science is being attempted in academia without
any formal training'' kicked off our discussion. The group brought expertise in
several different training models, from a two-day Software Carpentry workshop,
a three-week Open Science for Synthesis (OSS) program, to a semester-long
programs. We discussed the pros and cons of different training approaches, 
touching on informal learning, for example, where people learning the necessary
skills by asking questions of cross-disciplinary people (``boundary scientists'') in
their work environments.

Some papers explored gaps in training of early-career scientists. Industry
participants in our group confirmed this observation. We asked ourselves, ``Are traditional
courses failing?'' We think yes. Changes to undergraduate curricula requirements
are difficult. But as programming models become more complex, we have to raise
the skill levels. Skills must be improved not only in traditional programming --
learning languages and algorithms -- but also around professional software
development. We want to get to a point where ``Everyone has a new minimum now --
everyone knows Git.'' Developers also need training on licensing choices. Just
because a code is on Git does not mean that it is open.

``How do you get people to look for the training they need?'' wondered one
participant. People seek out opportunities like Software Carpentry to augment
skills. While some instructions can be done in institutional curricula,
independent groups (non-profits, institutes) have more flexibility. Some asked
whether Silicon Valley would be interested in funding training so that people are
better prepared to enter the workforce. Some felt that companies were reluctant
to deliver training for fear that their employees might then leave them. While others felt
that pushing all training to industry could lead to good technical people
``getting on the Google bus'' and leaving sciences.

``How will we know when people are trained effectively in these new skills?'' We
discussed certification. It can be hard to build the recognition of Java or
database certifications amongst all technologies. OSS offers badges to those
completing training. However we need to demonstrate proficiency, not just
completion. Google Summer of Code is a big CV augmentation for
participants---can we create something similar?

The group also discussed how to fund training. On participant observed that NYU
runs a six-week data-science training; companies grab the graduates and pay for
those they hire. OSS also used the NSF Software Institutes program as a vehicle
to fund training activities. Software Carpentry uses a collaborative teaching
approach where people publish open teaching materials and receive credits for
their use.

We then discussed career paths for those supporting sustainable software. While
tenure track is not the only option for students, the challenges for those who
remain in academia can be large. Research scientists are entirely dependent on
soft money, which can be unpredictable. Postdocs and those who do pursue tenure
track positions need to publish and see no rewards in software development.
These challenges were all identified at WSSSPE1. What actions would we recommend
to improve things? Altmetrics and download statistics may slowly change the
system and improve a developer's ability to receive credits for time invested in
software development. NSF's recognition of data products (including software) as
well as publications is also helping.

We asked ourselves, ``Where are there examples of things that are changing
because of this and how can we build momentum?'' One example demonstrates change
over time. In 2007, nanoHUB listed the academic reward structure as a problem in
an EDUCAUSE report where the authors note, ``In the future, nanoHUB researchers
are hoping to change the research culture. While they recognize that young
faculty members are unlikely to get tenure based on their nanoHUB contributions,
they hope to encourage faculty to think beyond their own research needs to
consider publishing tutorials and other content in their fields on the
nanoHUB site.'' Fast forward seven years later where, in a 2014 iSGTW article,
quote nanoHUB Principal Investigator (PI) Gerhard Klimeck, ``A former student of
mine published eight tools on nanoHUB, serving over 6,000 people with his tools.
He then joined a university as a professor and introduced nanoHUB. Use of the
gateway from that university skyrocketed; he used nanoHUB in existing classes,
created new classes, and infused it in his research. Ultimately, the professor's
department head associates his two-year rise to tenure with the notoriety and
innovation he gained through nanoHUB.''
